Analysis of Temporal Trends using Bayesian Generalized Additive Models

### 2025-04-25 analysis/code/010-summary-of-gams.qmd
working on GAMS explanation

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(brms)
library(tidyverse)
library(broom.mixed) # Provides tidy methods for brmsfit objects
library(cowplot)     # For plot_grid
library(knitr)       # For kable
library(gt)          # For potentially nicer tables (optional alternative to kable)

# Set brms options (optional)
# options(mc.cores = parallel::detectCores())

# Load your results object (replace with your actual loading code)
# load("path/to/your/results_brms.RData")
# For demonstration, let's assume 'results_brms' exists and contains valid models
# Filter out any models that failed to fit (are NULL)
results_brms_valid <- results_brms %>%
  filter(!sapply(mod_brms, is.null)) %>%
  filter(sapply(mod_brms, inherits, "brmsfit"))

# Check if any valid models remain
if(nrow(results_brms_valid) == 0) {
  stop("No valid brms models found in results_brms.")
}

@@

Separate GAMs were fitted for each response variable against publication year (`year`). The specific model structure for each variable was chosen based on the nature of the response data:

@@

@@

```{r check_divergences, echo=TRUE}
# Code to programmatically check divergences 
library(rstan)
sampler_params_list <- map(results_brms_valid$mod_brms, ~{
  if (!is.null(.x) && inherits(.x, "brmsfit")) {
    # Get parameters excluding warmup draws
    nuts_params(.x, inc_warmup = FALSE)
  } else {
    NULL
  }
})

@@

print("Number of Divergent Transitions per Model:")
print(divergence_summary)

@@

```{r convergence_summary_table, echo=FALSE}

@@

# --- Display the Table ---
# (kable formatting code remains the same)
kable(
  convergence_summary_df,
  caption = "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0 (e.g., < 1.01). Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0 for reliable results.",
  digits = c(NA, 3, 0, 0, 0), # Specify digits for numeric columns
  booktabs = TRUE
) %>%
  kableExtra::kable_styling(latex_options = "scale_down")

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(brms)
library(tidyverse)
library(broom.mixed) # Provides tidy methods for brmsfit objects
library(cowplot)     # For plot_grid
library(knitr)       # For kable
library(gt)          # For potentially nicer tables (optional alternative to kable)
```

@@

Separate GAMs were fitted for each response variable against publication year (`year`). The specific model structure for each variable was chosen based on the nature of the response data. @fig-brms-distros shows the distributions of the five variables related to scientific publications. The shape of these distributions informed our decisions about model specification. 
Recency of references: This variable, bounded between 0 and 1, exhibited a right-skewed distribution with notable occurrences at the boundaries (0 and 1). Consequently, a Zero-One Inflated Beta (zero_one_inflated_beta()) distribution family was selected. The associated logit link function is standard for modeling the mean (mu) of proportions, transforming the (0, 1) interval to the real line (-∞, +∞) and modeling the log-odds of the mean recency score for values strictly between 0 and 1.

Diversity of references & Relative title length (ln): Both variables displayed approximately symmetric, bell-shaped (normal) distributions. For these continuous variables, the Gaussian (gaussian()) family was employed with the canonical identity link function, meaning the linear predictor directly models the expected value (mean) of the response variable.

N. of authors & N. of pages: These count variables both exhibited highly right-skewed distributions. Therefore, the Negative Binomial (negbinomial()) family was chosen. The standard log link function was applied, ensuring that the predicted mean remains positive (μ = exp(η)) and implying that predictor effects are multiplicative on the original count scale, which is conventional for count data.

```{r}
#| label: fig-brms-distros
#| fig-cap: "Distributions of the response variables"

# prepare data for GAMS

# create data objects needed for this document
qmd_file <- here::here("analysis/paper/paper.qmd")

# A one-liner to run the R code in the main paper and create all the data objects
{ f <- tempfile(fileext = ".R"); on.exit(unlink(f, force = TRUE)); knitr::purl(qmd_file, output = f, documentation = 0, quiet = TRUE); source(f, local = FALSE); }

# on with the show
over_time <-  
  items_df %>%  
  left_join(items_df_title) %>%  
  left_join(shannon_per_item) %>%  
  filter(relative_title_length != -Inf, 
         relative_title_length !=  Inf,
         prices_index != "NaN" 
  ) %>% 
  mutate(log_authors_n = log(authors_n), 
         log_pages_n = log(pages_n), 
         journal_wrp = str_wrap(journal, 30)) %>%  
  select(year, 
         log_authors_n, 
         log_pages_n, 
         prices_index,  
         shannon, 
         relative_title_length,
         authors_n,
         pages_n)

over_time_long <-  
  over_time %>%  
  ungroup() %>%  
  select(-journal) %>%  
  gather(variable,  
         value, 
         -year) %>%  
  filter(value != -Inf, 
         value !=  Inf) %>%  
  mutate(variable = case_when( 
    variable == "prices_index"  ~ "Recency of references", 
    variable == "shannon"  ~ "Diversity of references", 
    variable == "relative_title_length"  ~ "Relative title length (ln)", 
    variable == "authors_n" ~ "N. of authors",
    variable == "pages_n" ~ "N. of pages"
  )) %>%  
  filter(!is.na(variable)) %>%  
  filter(!is.nan(value)) %>%  
  filter(!is.na(value)) %>%  
  filter(value != "NaN") %>%  
  mutate(value = parse_number(value))

# compute beta estimates so we can colour lines to indicate more or
# less hard

library(broom)

over_time_long_models <- 
  over_time_long %>%  
  group_nest(variable) %>%  
  mutate(model = map(data, ~tidy(lm(value ~ year, data = .)))) %>%  
  unnest(model) %>%  
  filter(term == 'year') %>%  
  mutate(becoming_more_scientific = case_when( 
    variable == "N. of authors"         & estimate > 0 ~ "TRUE", 
    variable == "N. of pages"           & estimate < 0 ~ "TRUE", 
    variable == "N. of refs (sqrt)"          & estimate < 0 ~ "TRUE", 
    variable == "Recency of references"      & estimate > 0 ~ "TRUE", 
    variable == "Relative title length (ln)" & estimate > 0 ~ "TRUE", 
    variable == "Diversity of references"    & estimate < 0 ~ "TRUE",
    TRUE ~ "FALSE"
  ))

# join with data
over_time_long_colour <- 
  over_time_long %>% 
  left_join(over_time_long_models)

over_time_long_colour_gams_testing_nested_df <- 
  over_time_long_colour %>% 
  nest(.by = variable)

# check the distributions of each variable
each_variable <- 
map(over_time_long_colour_gams_testing_nested_df$data, 
    ~.x$value)

names(each_variable) <- over_time_long_colour_gams_testing_nested_df$variable

stack(each_variable) %>% 
  ggplot() +
  aes(values) +
  geom_histogram() +
  facet_wrap(~ind, scales = "free") +
  theme_minimal()
```

The table below summarises the selections made for each variable, based on their distributions obsvered in @fig-brms-distros.

@@

```{r}
#| eval: false

# ---  fit models with brms -------------------------------
# this takes a few hours

library(brms)
library(dplyr)
library(purrr)

# --- Set options for brms (optional, can help speed up) ---
options(mc.cores = parallel::detectCores()) # Use all available CPU cores

# --- the nested df is 'over_time_long_colour_gams_testing_nested_df' 
# with list-column 'data' and character column 'variable'

results_brms <- over_time_long_colour_gams_testing_nested_df %>%
  mutate(
    # Create a new column for brms models to avoid overwriting gam models if needed
    mod_brms = map2(data, variable, ~{
      current_df <- .x        # The dataframe for the current row
      current_var <- .y       # The variable name for the current row
      response_col <- "value" #  this matches the response column name in nested dfs

      message(paste("Fitting brms model for:", current_var)) # Progress message

      # --- 1. Choose the brms family based on the variable name ---
      chosen_family <- if (current_var == "N. of authors") {
        negbinomial() # For count data
      } else if (current_var == "N. of pages") {
        negbinomial() # For count data
      } else if (current_var == "Recency of references") {
        zero_one_inflated_beta() # For 0/1 inflated proportions
      } else if (current_var == "Diversity of references") {
        gaussian() # For continuous, approx normal data
      } else if (current_var == "Relative title length (ln)") {
        gaussian() # For continuous, normal data
      } else {
        warning(paste("Unknown variable:", current_var, "- using Gaussian default."))
        gaussian() # Default fallback
      }

      # --- 2. Construct the formula using brms::bf() ---
      # Handle the special structure for Zero-One Inflated Beta
      if (inherits(chosen_family, "zero_one_inflated_beta")) {
        # Formula for the mean (between 0 and 1), plus formulas for
        # zoi (zero-inflation probability) and coi (one-inflation probability).
        # zoi and coi are modelled simply as intercepts.
        brm_formula <- bf(
          paste(response_col, "~ s(year, bs = 'cr')"), 
          # Smooth term for mean
          zoi ~ 1,                                    
          # Intercept for prob of 0
          coi ~ 1                                     
          # Intercept for prob of 1
        )
      } else {
        # Standard formula for other families
        brm_formula <- bf(
          paste(response_col, "~ s(year, bs = 'cr')")
        )
      }

      # --- 3. Fit the brms model ---
      # Using tryCatch to prevent one failure from stopping everything
      tryCatch({
        brm(
          formula = brm_formula,
          family = chosen_family,
          data = current_df,
          # --- Stan control arguments (adjust as needed) ---
          cores = 4,  # Number of cores for parallel chains 
          chains = 4, # Number of Markov chains (4 is standard for final runs)
          iter = 3000, # Total iterations per chain (includes warmup)
          warmup = 1000,# Number of warmup iterations (discarded)
          control = list(adapt_delta = 0.9999), # for helping convergence for complex models
          seed = 123, # For reproducibility
          silent = 2, # Suppress compilation messages from Stan
          refresh = 0 # Suppress iteration progress updates

        )
      }, error = function(e) {
        # Print a warning if a model fails to fit
        warning(paste("brms fitting failed for variable:", 
                      current_var, "\nError:", e$message))
        return(NULL) # Return NULL for failed models
      })
    }) # End of map2
  ) # End of mutate

# quite a large file
saveRDS(results_brms, here::here("analysis/data/results_brms.RData"))

```{r}

# Load my brms results object so we don't have to re-run the above block
load(here::here("analysis/data/results_brms.RData"))

# For demonstration, let's assume 'results_brms' exists and contains valid models
# Filter out any models that failed to fit (are NULL)
results_brms_valid <- results_brms %>%
  filter(!sapply(mod_brms, is.null)) %>%
  filter(sapply(mod_brms, inherits, "brmsfit"))

@@

```{r check_divergences, echo=TRUE}
#| tbl-cap: "Number of divergences for each variable"

# Code to programmatically check divergences 
library(rstan)
sampler_params_list <- map(results_brms_valid$mod_brms, ~{
  if (!is.null(.x) && inherits(.x, "brmsfit")) {
    # Get parameters excluding warmup draws
    nuts_params(.x, inc_warmup = FALSE)
  } else {
    NULL
  }
})

@@

kable(divergence_summary)

@@

```{r convergence_summary_table, echo=FALSE}
#| tbl-cap: "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0 (e.g., < 1.01). Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0 for reliable results."

@@

# --- Display the Table ---
# (kable formatting code remains the same)
kable(
  convergence_summary_df,
  caption = ,
  digits = c(NA, 3, 0, 0, 0), # Specify digits for numeric columns
  booktabs = TRUE
) %>%
  kableExtra::kable_styling(latex_options = "scale_down")
